---
title: "Emulating the natural variables"
output: html_notebook
---

Glen Pridham 2024.

This file describes how to convert a set of biomarkers into natural variables. This file does not consider sex-specific effects, and as a result is not completely accurate (e.g. R2 reaches its max of .97 for z1, rather than 1). The errors are small so I'm keeping this file, but you should use the sex-specific one.



The biomarkers used are:
albumin in g/L
calcium in mmol/L
chloride in mmol/L
glucose (random) in mmol/L
hemoglobin in g/L
magnesium in mmol/L
phosphate in mmol/L
platelet count in billion/L
potassium in mmol/L
creatinine in mumol/L (micro mols per liter)
urea in mmol/L
total CO2 in mmol/L
white blood cell count in billion/L
*Note: these are all measured BEFORE the dialysis session.

There are three cases discussed here:
1. you have all 14 biomarkers,
2. you have a subset of the 14,
3. you have a subset of the 6 validation set (creatinine, albumin, urea, phosphate, calcium, total_co2).


put the location of your files here (i.e. where the coefficients are saved)
```{r}
fileDir = "results" # put your file directory name here e.g. C:/glen/results
```

#########################################################
########## Case 1 - you have all 14 biomarkers ########## 
#########################################################

First you'll need to load in your data. I'll just generate some fake (unrealistic) data.
```{r}
vars = c("pre.creatinine", "platelets", "albumin", "pre.urea",
          "hemoglobin", "sodium", "chloride", "wbc", "potassium", "phosphate",     
          "glucose.r", "magnesium", "total.co2", "calcium")    

data = data.frame(albumin=rnorm(100))
for (j in 1:length(vars)) data[,vars[j]] = rnorm(100)
print(head(data))
```

Next we must convert from the original units to the scaled units. 
```{r}
data_sd = read.csv(sprintf("%s/kidney_data_sd.csv",fileDir),row.names=1)  
mu_sex  = read.csv(sprintf("%s/kidney_data_mu.csv",fileDir),row.names=1) #sex 0: male, sex 1: female
``` 

To scale we have to subtract off the sex-specific mean then divide by the sd (standard deviation).

The scaling is done by sex, so we'll need to know the sex of each individual. For simplicity let's suppose the first 50 are males and the second are females
```{r}
males = data[1:50,]
females = data[51:100,]

for (j in 1:ncol(males))
{
  males[,j] = (males[,j]-mu_sex[1,colnames(males)[j]])/data_sd[colnames(males)[j],1]
}

for (j in 1:ncol(females))
{
  females[,j] = (females[,j]-mu_sex[2,colnames(females)[j]])/data_sd[colnames(females)[j],1]
}
```


Since we have all 14 biomarkers we can use the full transformation. The real and imaginary parts are saved separately and need to be combined (and converted to a single matrix).
```{r}
rePinv = read.csv(sprintf("%s/natural_variable_transformation_Pinv_real.csv",fileDir),row.names=1)
imPinv = read.csv(sprintf("%s/natural_variable_transformation_Pinv_imaginary.csv",fileDir),row.names = 1)
Pinv = as.matrix(rePinv)+1.i*as.matrix(imPinv)

print(Pinv)
```
Now we transform. The formal is Pinv%*%x, where x is a vector for each individual. This can be done all at once using
```{r}
zm = as.matrix(males[,colnames(Pinv)])%*%t(Pinv)
zf = as.matrix(females[,colnames(Pinv)])%*%t(Pinv)
head(zm)
```
Now you have all of the natural variables.

You'll notice that some are complex. In fact, z02/z03 and z06/z07 are complex conjugates. This means we can alternatively deal with entirely real z using:
```{r}
zm[,3] = Im(zm[,2])
zf[,3] = Im(zf[,2])
colnames(zm)[3] = "Im(z02)"
colnames(zf)[3] = "Im(z02)"
zm[,7] = Im(zm[,6])
zf[,7] = Im(zf[,6])
colnames(zm)[7] = "Im(z06)"
colnames(zf)[7] = "Im(z06)"

zm = Re(zm)
zf = Re(zf)

print(head(zm))
```

You now have all of the natural variables used in the paper.


#################################################################
########## Case 2 - you have SOME of the 14 biomarkers ########## 
#################################################################

First you'll need to load in your data. I'll generate some (limited) fake data.
```{r}
data = data.frame(albumin=rnorm(100),pre_creatinine=rnorm(100),platelets=rnorm(100),calcium=rnorm(100))

print(head(data))
```

Let's look at the emulator coefficients, for example z01
```{r}

C1 = read.csv(sprintf("%s/kidney_emulator_models_z01_coefficients.csv",fileDir))

print(C1)
```

Each row is a different emulator model. The columns are sorted by the best predictors, *which differs for each z*. For the case of z01 those are creatinine then platelets then albumin ...

Each model has an associated R2 (coefficient of variation) value for the strength of the approximation. Pick the best model for which you have ALL non-zero variables. We have pre_creatinine, platelets and albumin, so the best model is number 3:
```{r}
C1[3,]
```

Now we transform.
```{r}
z1 = C1[3,"X.Intercept."] + as.matrix(data[,c("pre_creatinine","platelets","albumin")])%*%t(C1[3,c("pre_creatinine","platelets","albumin")])
head(z1)
```
That's it. Now you repeat for all 14 zs. Note that z3 = Im(z2) and z7 = Im(z6), as in the paper.


###########################################################################
########## Case 3 - you have some of the 6 validation biomarkers ########## 
###########################################################################

In some cases your biomarker set might better align with the  validation set, which are:
creatinine, albumin, urea, phosphate, calcium, and total_co2.

If this is the case, you should instead use the "convenience" model. This just repeats case 2 but using a different set of coefficients.

First you'll need to load in your data. I'll generate some (limited) fake data.
```{r}
data = data.frame(albumin=rnorm(100),pre_creatinine=rnorm(100),pre_urea=rnorm(100))

print(head(data))
```

Let's look at the emulator coefficients, for example z01
```{r}

C1 = read.csv(sprintf("%s/kidney_emulator_models_z01_coefficients.csv",fileDir))

print(C1)
```

Each row is a different emulator model. The columns are sorted by the best predictors, *which differs for each z*. For the case of z01 those are creatinine then platelets then albumin ...

Each model has an associated R2 (coefficient of variation) value for the strength of the approximation. Pick the best model for which you have ALL non-zero variables. We have just pre_creatinine, so the best model is number 1:
```{r}
C1[1,]
```

Now we transform.
```{r}
z1 = C1[1,"X.Intercept."] + data[,"pre_creatinine"]*C1[1,"pre_creatinine"]
head(z1)
```
 Done